---
title: "Non-Negative Matrix Factorization (NMF) report"
author: "Health Data Science Unit"
date: "`r date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    toc_depth: 4
    number_sections: false
params:
  assayID:   "foo"
  annot_tumor:  "path/to/metadata"
  annot_cells:  "path/to/metadata"
  nmf_kmin:  2
  nmf_kmax:  6
  nmf_iter:  10
  nmf:       "path/to/save"
  norm_nmfW: "path/to/save"
  norm_nmfH: "path/to/save"
  matrix:    "path/to/matrix"
  K         : 4
  hmatrix_wnorm : "path/to/matrix"
  wmatrix_wnorm : "path/to/matrix"
  nmf_features  : "path/to/matrix"
  hmatrix_hnorm : "path/to/matrix"
editor_options: 
  chunk_output_type: console
---


```{r setup, include=FALSE}
#this chunk sets global options for Rmarkdown
knitr::opts_chunk$set(echo = FALSE)
#setting working directory. Should be home directory of the project
options(knitr.table.format="html")

library(Bratwurst)
library(viridis)
library(ComplexHeatmap)
library(circlize)
library(ggplot2)
library(knitr)
library(tidyverse)



```

# `r params$assayID`

## Data loading
  
Read Non Negative matrix
  
```{r dataloading, results="asis", cache=FALSE, message=FALSE}
##––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––##
##                         Read normalized data                               ##
##––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––##
# read normalized matrix
norm.mat <- readRDS(params$matrix)

# read metadata
annot_tumor <- readRDS(params$annot_tumor)
annot_cells <- readRDS(params$annot_cells)

metadata <- bind_rows(annot_tumor, annot_cells)
metadata <- metadata[metadata$avail.ChIPseq, ]
metadata  <- metadata[match(colnames(norm.mat), metadata$ProjectID),]

##––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––##
##                          Print dataset dimension                           ##
##––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––##
tmp.dim <- data.frame(dim(norm.mat))
colnames(tmp.dim) <- "Value"

cat("Dimension of input Matrix  \n  ") 
kable(tmp.dim) 

cat("Removing rows with only 0 values  \n  ") 
norm.mat <- norm.mat[!rowSums(norm.mat) == 0,,drop=FALSE]
cat("Removing columns with only 0 values  \n  ") 
norm.mat <- norm.mat[,!colSums(norm.mat) == 0,drop=FALSE]

cat("Dimension of input Matrix  \n  ") 
tmp.dim <- data.frame(dim(norm.mat))
colnames(tmp.dim) <- "Value"
kable(tmp.dim) 




```
  

# Applying NMF

Applying Non-Negative Matrix Factorization (NMF) to normalized transcriptome data (RNAseq) 
  
  
```{r NMF_param, echo=TRUE, warning=FALSE}
##––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––##
##              Parameters to run NMF in GPUs using  pythonCuda               ##
##––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––##
k.min <- params$nmf_kmin
k.max <- params$nmf_kmax
outer.iter <- params$nmf_iter
inner.iter <- 2*10^4

# path to save NMF results
path.NMF.run       <- params$nmf
path.normNMF.run   <- params$norm_nmfW
path.normNMF_H.run <- params$norm_nmfH

#create directory to save results
#dir.create("NMF", recursive = TRUE)

```


```{r NMF_run, eval=TRUE, message=FALSE}
##––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––##
##              Create nmf experiment object and run NMF                      ##
##––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––##
nmf.exp <- nmfExperimentFromMatrix(matrix = norm.mat)


nmf.exp <- runNMFtensor(nmf.exp, 
                        k.min = k.min, 
                        k.max = k.max, 
                        outer.iter = outer.iter, 
                        inner.iter = inner.iter, 
                        conver.test.stop.threshold = 1000)
# save results
saveRDS(nmf.exp, file = path.NMF.run)

##––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––##
##                             Normalize NMF                                  ##
##––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––##
# normalize NMF
norm.nmf.exp <- normalizeW(nmf.exp)
norm.nmf.exp.H <- normalizeH(nmf.exp)

##––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––##
##                    K stats and normalization                               ##
##––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––##
## Estimate K stats
my.kstats <- function(NMFexperiment){
  # calc different k stats
  NMFexperiment <- computeFrobErrorStats(NMFexperiment)
  NMFexperiment <- computeSilhoutteWidth(NMFexperiment)
  NMFexperiment <- computeCopheneticCoeff(NMFexperiment)
  NMFexperiment <- computeAmariDistances(NMFexperiment)
  return(NMFexperiment)
}
norm.nmf.exp <- my.kstats(norm.nmf.exp)

##––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––##
##                            Save results                                    ##
##––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––##
#save results as RDS file
saveRDS(norm.nmf.exp, file = path.normNMF.run)
saveRDS(norm.nmf.exp.H, file = path.normNMF_H.run)

```
 
 
### Factorization quality metrics and optimal K

Based on the results of the factorization quality metrics, an optimal number of signatures (k) must be chosen:

  
```{r NMF_optK, results='hide',fig.keep='all', eval=TRUE, warning=FALSE, message=FALSE}
##––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––##
##                            Plot K stats                                    ##
##––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––##
my.plotKstats <- function(NMFexperiment, title){
  # visualize k stats
  gg.optKr <- plotKStats(NMFexperiment)
  gg.optKr <- gg.optKr + theme_bw() + 
    ggtitle(title) +
    theme(plot.title=element_text(hjust=0.5))
  return(gg.optKr)
}

gg.optKr <- my.plotKstats(norm.nmf.exp, "NMF factorization quality metrics")
gg.optKr


##––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––##
##                        Generate river plot                                 ##
##––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––##
river <- generateRiverplot(norm.nmf.exp)
plot(river, plot_area=1, yscale=0.6, nodewidth=0.5)

# pdf(file = paste0(figs.path, "/NMF_riverplot.pdf"), width=8, height=6)
# plot(river, plot_area=1, yscale=0.6, nodewidth=0.5)
# dev.off()



```

Minize the Frobenius error, the coefficient of variation and the mean Amari distance, while maximizing the sum and mean silhouette width and the cophenic coefficient.
  
### H Matrix, W normalized:  {.tabset}  
  
  
```{r Hmatrix_Wnorm, fig.width=10, fig.height=5.5, out.width="90%", results='asis', eval=TRUE, warning=FALSE}

#Annotation for H matrix heatmap
typeIDs <- unique(as.character(metadata$Type))
type.colVector <- list(Type = setNames(inferno(n=length(typeIDs)), typeIDs))
heat.anno <- HeatmapAnnotation(df = data.frame(Type = metadata$Type),
                               col = type.colVector,
                               show_annotation_name = TRUE, na_col = "white")

##––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––##
##              Generate H matrix heatmap, W normalized                       ##
##––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––##


for(ki in names(norm.nmf.exp@HMatrixList)) {
  cat("\n")
  cat("  \n#### H matrix for k=",  ki, "  \n  ")
  #plot H matrix
  tmp.hmatrix <- HMatrix(norm.nmf.exp, k = ki)
  colnames(tmp.hmatrix) <- colnames(norm.nmf.exp)
  h.heatmap <- Heatmap(tmp.hmatrix,
                       col = colorRamp2(seq(min(tmp.hmatrix), max(tmp.hmatrix), length.out = 100), viridis(n=100)),
                       name = "Exposure",
                       clustering_distance_columns = 'pearson',
                       show_column_dend = FALSE,
                       heatmap_legend_param = 
                         list(color_bar = "continuous", legend_height=unit(2, "cm")),
                       top_annotation = heat.anno,
                       show_column_names = FALSE,
                       show_row_names = FALSE,
                       cluster_rows = FALSE)
  print(h.heatmap)
  # h.heatmap.list[[as.character(ki)]] <- h.heatmap
}


```



# Save NMF H and W matrices




```{r saveNMFobjects}
##––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––##
##               Select opt K and compute feature stats                     ##
##––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––##
signature_idx <- setNames(1:params$K, paste0("Sig.", 1:params$K))

K <- params$K
norm.nmf.exp <- setOptK(norm.nmf.exp, K)
norm.nmf.exp@OptK

##––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––##
##                             NMH H Matrix                                 ##
##––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––##
norm.nmf.hmatrix <- HMatrix(norm.nmf.exp, k = K)[signature_idx,]
colnames(norm.nmf.hmatrix) <- colnames(norm.nmf.exp)
rownames(norm.nmf.hmatrix) <- names(signature_idx)
saveRDS(norm.nmf.hmatrix, params$hmatrix_wnorm)

##––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––##
##                             NMH W Matrix                                 ##
##––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––##
norm.nmf.wmatrix <- WMatrix(norm.nmf.exp, k = K)[,signature_idx]
colnames(norm.nmf.wmatrix) <- names(signature_idx)
rownames(norm.nmf.wmatrix) <- rownames(norm.nmf.exp)
saveRDS(norm.nmf.wmatrix, params$wmatrix_wnorm)

##––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––##
##                      NMF Signature features                              ##
##––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––##
norm.nmf.exp <- computeFeatureStats(norm.nmf.exp)
norm.nmf.exp <- computeSignatureSpecificFeatures(norm.nmf.exp, 1)
SignaFeat <- do.call(cbind, lapply(1:norm.nmf.exp@OptK, function(i){
  as.numeric(substr(norm.nmf.exp@FeatureStats$cluster, i, i))
}))
SignaFeat <- as.data.frame(SignaFeat)[,signature_idx]
#rownames are features names
rownames(SignaFeat) <- rownames(norm.nmf.exp)
#colnames are signatures
colnames(SignaFeat) <- names(signature_idx)
saveRDS(SignaFeat, params$nmf_features)


```


```{r saveNMFobjectsHnorm}
##––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––##
##                             NMH H Matrix                                 ##
##––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––##
norm.nmf.hmatrix.hnorm <- HMatrix(norm.nmf.exp.H, k = K)[signature_idx,]
colnames(norm.nmf.hmatrix.hnorm) <- colnames(norm.nmf.exp.H)
rownames(norm.nmf.hmatrix.hnorm) <- names(signature_idx)
saveRDS(norm.nmf.hmatrix, params$hmatrix_hnorm)

```




