---
title: "Non-Negative Matrix Factorization (NMF) report"
author: "Health Data Science Unit"
date: "`r date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    toc_depth: 4
    number_sections: false
params:
  assayID:   "foo"
  work_dir:  "foo"
  nmf_kmin:  2
  nmf_kmax:  6
  nmf_iter:  10
  nmf:       "path/to/save"
  norm_nmfW: "path/to/save"
  norm_nmfH: "path/to/save"
  matrix:    "path/to/matrix"
  metadata:  "path/to/metadata"
  
---


```{r setup, include=FALSE}
#this chunk sets global options for Rmarkdown
knitr::opts_chunk$set(echo = FALSE)
#setting working directory. Should be home directory of the project
knitr::opts_knit$set(root.dir=params$work_dir)
options(knitr.table.format="html")

library(Bratwurst)
library(viridis)
library(ComplexHeatmap)
library(circlize)
library(ggplot2)
library(knitr)



```

# `r params$assayID`

## Data loading
  
Read Non Negative matrix
  
```{r dataloading, results="asis", cache=FALSE, message=FALSE}
##––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––##
##                         Read normalized data                               ##
##––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––##
# read normalized matrix
norm.mat <- readRDS(params$matrix)

# read metadata
metadata <- readRDS(params$metadata)

##––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––##
##                          Print dataset dimension                           ##
##––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––##
tmp.dim <- data.frame(dim(norm.mat))
colnames(tmp.dim) <- "Value"

cat("Dimension of input Matrix  \n  ") 
kable(tmp.dim) 

cat("Removing rows with only 0 values  \n  ") 
norm.mat <- norm.mat[!rowSums(norm.mat) == 0,,drop=FALSE]
cat("Removing columns with only 0 values  \n  ") 
norm.mat <- norm.mat[,!colSums(norm.mat) == 0,drop=FALSE]

cat("Dimension of input Matrix  \n  ") 
tmp.dim <- data.frame(dim(norm.mat))
colnames(tmp.dim) <- "Value"
kable(tmp.dim) 




```
  

# Applying NMF

Applying Non-Negative Matrix Factorization (NMF) to normalized transcriptome data (RNAseq) 
  
  
```{r NMF_param, echo=TRUE, warning=FALSE}
##––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––##
##              Parameters to run NMF in GPUs using  pythonCuda               ##
##––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––##
k.min <- params$nmf_kmin
k.max <- params$nmf_kmax
outer.iter <- params$nmf_iter
inner.iter <- 2*10^4

# path to save NMF results
path.NMF.run       <- params$nmf_kmin
path.normNMF.run   <- params$norm_nmfW
path.normNMF_H.run <- params$norm_nmfH

#create directory to save results
#dir.create("NMF", recursive = TRUE)

```


```{r NMF_pyCuda, eval=TRUE, message=FALSE}
##––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––##
##                        Create nmf experiment object                        ##
##––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––##
nmf.exp <- nmfExperimentFromMatrix(matrix = norm.mat)

##––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––##
##                                   Run NMF                                ##
##––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––##
nmf.exp <- runNmfGpuPyCuda(
  nmf.exp = nmf.exp,
  k.min = k.min,
  k.max = k.max,
  outer.iter = outer.iter,
  inner.iter = inner.iter,
  tmp.path = paste0(params$work_dir, "/tmp"),
  seed = TRUE,
  gpu.id = 0)

# save results
saveRDS(nmf.exp, file = path.NMF.run)

##––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––##
##                             Normalize NMF                                ##
##––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––##
# normalize NMF
norm.nmf.exp <- normalizeW(nmf.exp)
norm.nmf.exp.H <- normalizeH(nmf.exp)

##––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––##
##                    K stats and normalization                             ##
##––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––##
## Estimate K stats
my.kstats <- function(NMFexperiment){
  # calc different k stats
  NMFexperiment <- computeFrobErrorStats(NMFexperiment)
  NMFexperiment <- computeSilhoutteWidth(NMFexperiment)
  NMFexperiment <- computeCopheneticCoeff(NMFexperiment)
  NMFexperiment <- computeAmariDistances(NMFexperiment)
  return(NMFexperiment)
}
norm.nmf.exp <- my.kstats(norm.nmf.exp)

##––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––##
##                            Save results                                  ##
##––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––##
#save results as RDS file
saveRDS(norm.nmf.exp, file = path.normNMF.run)
saveRDS(norm.nmf.exp.H, file = path.normNMF_H.run)

```
 
 
### Factorization quality metrics and optimal K

Based on the results of the factorization quality metrics, an optimal number of signatures (k) must be chosen:

```{r fixBratwurst}
```

  
```{r NMF_optK, results='hide',fig.keep='all', eval=TRUE, warning=FALSE, message=FALSE}
##––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––##
##                            Plot K stats                                    ##
##––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––##
my.plotKstats <- function(NMFexperiment, title){
  # visualize k stats
  gg.optKr <- plotKStats(NMFexperiment)
  gg.optKr <- gg.optKr + theme_bw() + 
    ggtitle(title) +
    theme(plot.title=element_text(hjust=0.5))
  return(gg.optKr)
}

gg.optKr <- my.plotKstats(norm.nmf.exp, "NMF factorization quality metrics")
gg.optKr

# Path to save results
figs.path <- file.path(params$work_dir, "/figs")

#create directory to save plots
dir.create(figs.path, recursive = TRUE)

#save K stats
pdf(file = paste0(figs.path, "/NMF_factorization.metrics.pdf"), width=8, height=6)
gg.optKr
dev.off()

##––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––##
##                        Generate river plot                                 ##
##––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––##
#river <- generateRiverplot(norm.nmf.exp)
river <- my_generateRiverplot(norm.nmf.exp)
plot(river, plot_area=1, yscale=0.6, nodewidth=0.5)

pdf(file = paste0(figs.path, "/NMF_riverplot.pdf"), width=8, height=6)
plot(river, plot_area=1, yscale=0.6, nodewidth=0.5)
dev.off()



```

Minize the Frobenius error, the coefficient of variation and the mean Amari distance, while maximizing the sum and mean silhouette width and the cophenic coefficient.
  
### H Matrix, W normalized:  {.tabset}  
  
  
```{r Hmatrix_Wnorm, fig.width=10, fig.height=5.5, out.width="90%", results='asis', eval=TRUE, warning=FALSE}
##––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––##
##                        H matrix heatmap annotation                         ##
##––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––##
# #Annotation for H matrix heatmap
# typeIDs <- unique(metadata$type)
# type.colVector <- list(Type       = structure(inferno(n=length(typeIDs)), names=typeIDs),
#                        G1_score   = colorRamp2(seq(0, 1, length.out = 100), plasma(n=100, end =.8)),
#                        G2M_score  = colorRamp2(seq(0, 1, length.out = 100), plasma(n=100, end =.8)))
# heat.anno <- HeatmapAnnotation(df = data.frame(Type      = metadata$type,
#                                                G1_score  = cyclone$scores$G1,
#                                                G2M_score = cyclone$scores$G2M),
#                                col = type.colVector,
#                                show_annotation_name = TRUE, na_col = "white",
#                                annotation_legend_param=list(
#                                  G1_score=list(color_bar="continuous"),
#                                  G2M_score=list(color_bar="continuous")))


#Annotation for H matrix heatmap
typeIDs <- unique(metadata$type)
type.colVector <- list(Type = setNames(inferno(n=length(typeIDs)), typeIDs))
heat.anno <- HeatmapAnnotation(df = data.frame(Type = metadata$type),
                               col = type.colVector,
                               show_annotation_name = TRUE, na_col = "white")

##––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––##
##              Generate H matrix heatmap, W normalized                       ##
##––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––##

h.heatmap.list <- list()
for(ki in names(norm.nmf.exp@HMatrixList)) {
  cat("\n")
  cat("  \n#### H matrix for k=",  ki, "  \n  ")
  #plot H matrix
  tmp.hmatrix <- HMatrix(norm.nmf.exp, k = ki)
  colnames(tmp.hmatrix) <- colnames(norm.nmf.exp)
  h.heatmap <- Heatmap(tmp.hmatrix,
                       col = colorRamp2(seq(min(tmp.hmatrix), max(tmp.hmatrix), length.out = 100), viridis(n=100)),
                       name = "Exposure",
                       clustering_distance_columns = 'pearson',
                       show_column_dend = FALSE,
                       heatmap_legend_param = 
                         list(color_bar = "continuous", legend_height=unit(2, "cm")),
                       top_annotation = heat.anno,
                       show_column_names = FALSE,
                       show_row_names = FALSE,
                       cluster_rows = FALSE)
  print(h.heatmap)
  h.heatmap.list[[as.character(ki)]] <- h.heatmap
}

#save H matrix heatmap
pdf(file = paste0(figs.path, "/NMF_Hmatrix_wnorm_heatmap.pdf"), width=10, height=7)
for(ki in names(h.heatmap.list)) {
  print(h.heatmap.list[[as.character(ki)]])
}
dev.off()


```


### H Matrix, H normalized:  {.tabset}  
  
  
```{r Hmatrix_Hnorm, fig.width=10, fig.height=5.5, out.width="90%", results='asis', eval=TRUE, warning=FALSE}

##––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––##
##              Generate H matrix heatmap, H normalized                       ##
##––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––##

h.heatmap.list <- list()
for(ki in names(norm.nmf.exp.H@HMatrixList)) {
  cat("\n")
  cat("  \n#### H matrix for k=",  ki, "  \n  ")
  #plot H matrix
  tmp.hmatrix <- HMatrix(norm.nmf.exp.H, k = ki)
  colnames(tmp.hmatrix) <- colnames(norm.nmf.exp.H)
  h.heatmap <- Heatmap(tmp.hmatrix,
                       col = colorRamp2(seq(min(tmp.hmatrix), max(tmp.hmatrix), length.out = 100), viridis(n=100)),
                       name = "Exposure",
                       clustering_distance_columns = 'pearson',
                       show_column_dend = FALSE,
                       heatmap_legend_param = 
                         list(color_bar = "continuous", legend_height=unit(2, "cm")),
                       top_annotation = heat.anno,
                       show_column_names = FALSE,
                       show_row_names = FALSE,
                       cluster_rows = FALSE)
  print(h.heatmap)
  h.heatmap.list[[as.character(ki)]] <- h.heatmap
}

#save H matrix heatmap
pdf(file = paste0(figs.path, "/NMF_Hmatrix_hnorm_heatmap.pdf"), width=10, height=7)
for(ki in names(h.heatmap.list)) {
  print(h.heatmap.list[[as.character(ki)]])
}
dev.off()

```

### Save NMF objects


```{r saveNMFobjects, eval=FALSE, warning=FALSE}
path_results_NMF <- file.path(params$work_dir, "NMF/Wnormalized")

for(K in k.min:k.max) {
  ##––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––##
  ##               Select opt K and compute feature stats                     ##
  ##––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––##
  path_results_K <- paste0(path_results_NMF, "/K", K)
  dir.create(path_results_K, recursive = TRUE)
  
  norm.nmf.exp <- setOptK(norm.nmf.exp, K)
  norm.nmf.exp@OptK
  
  ##––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––##
  ##                             NMH H Matrix                                 ##
  ##––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––##
  norm.nmf.hmatrix <- HMatrix(norm.nmf.exp, k = K)
  colnames(norm.nmf.hmatrix) <- colnames(norm.nmf.exp)
  rownames(norm.nmf.hmatrix) <- paste0("Signature", 1:K)
  saveRDS(norm.nmf.hmatrix, paste0(path_results_K, "/NMF_hmatrix.K", K, ".RDS"))
  
  ##––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––##
  ##                             NMH W Matrix                                 ##
  ##––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––##
  norm.nmf.wmatrix <- WMatrix(norm.nmf.exp, k = K)
  colnames(norm.nmf.wmatrix) <- paste0("Signature", 1:K)
  rownames(norm.nmf.wmatrix) <- rownames(norm.nmf.exp)
  saveRDS(norm.nmf.wmatrix, paste0(path_results_K, "/NMF_wmatrix.K", K, ".RDS"))
  
  ##––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––##
  ##                          NMH W Matrix Z scores                           ##
  ##––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––##
  norm.nmf.wmatrix.zscores <- apply(norm.nmf.wmatrix, MARGIN=2, function(wmat_score){
    (wmat_score - median(wmat_score)) / mad(wmat_score)
  })
  colnames(norm.nmf.wmatrix.zscores) <- paste0("Zscore_", colnames(norm.nmf.wmatrix.zscores))
  saveRDS(norm.nmf.wmatrix.zscores, paste0(path_results_K, "/NMF_wmatrix-Zscore.K", K, ".RDS"))
  
  ##––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––##
  ##                      NMF Signature features                              ##
  ##––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––##
  if (K > 2) {
    norm.nmf.exp <- computeFeatureStats(norm.nmf.exp)
    norm.nmf.exp <- computeSignatureSpecificFeatures(norm.nmf.exp, 1)
    SignaFeat <- do.call(cbind, lapply(1:norm.nmf.exp@OptK, function(i){
      as.numeric(substr(norm.nmf.exp@FeatureStats$cluster, i, i))
    }))
    SignaFeat <- as.data.frame(SignaFeat)
    #rownames are features names
    rownames(SignaFeat) <- rownames(norm.nmf.exp)
    #conames are signatures
    colnames(SignaFeat) <- paste0("Signature", 1:norm.nmf.exp@OptK)
    saveRDS(SignaFeat, paste0(path_results_K, "/NMF_Features.K", K, ".RDS"))
  }
}



```


```{r saveNMFobjectsHnorm, eval=FALSE}
path_results_NMF <- file.path(params$work_dir, "NMF/Hnormalized")

for(K in k.min:k.max) {
  ##––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––##
  ##               Select opt K and compute feature stats                     ##
  ##––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––##
  path_results_K <- paste0(path_results_NMF, "/K", K)
  dir.create(path_results_K, recursive = TRUE)
  
  norm.nmf.exp.H <- setOptK(norm.nmf.exp.H, K)
  norm.nmf.exp.H@OptK
  sig.names <- paste0("Signature", 1:K)
  
  ##––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––##
  ##                             NMH H Matrix                                 ##
  ##––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––––##
  norm.nmf.hmatrix <- HMatrix(norm.nmf.exp.H, k = K)
  colnames(norm.nmf.hmatrix) <- colnames(norm.nmf.exp.H)
  rownames(norm.nmf.hmatrix) <- sig.names
  saveRDS(norm.nmf.hmatrix, paste0(path_results_K, "/NMF_hmatrix.K", K, ".RDS"))
}




# remove tmp dir
unlink(file.path(params$work_dir, "tmp"), recursive=TRUE)
```


```{r fixBratwurst}
my_generateRiverplot <- function(nmf.exp, edges.cutoff = 0, useH=FALSE,
                                color=TRUE) {
    if(useH)
      W.list <- lapply(HMatrix(nmf.exp), t)
    else
      W.list <- WMatrix(nmf.exp)
    rank.max <- max(as.numeric(names(W.list)))
    rank.min <- min(as.numeric(names(W.list)))
    nodes <- do.call(rbind, lapply(rank.min:rank.max, function(i) {
      data.frame(ID = sapply(1:i, function(n) paste(i, "_S", n, sep = "")),
                 x = i)
    }))
    edges <- do.call(rbind, lapply(rank.min:(rank.max - 1), function(i){
      df <- data.frame(
        N1 = rep(sapply(1:i, function(n) paste(i, "_S", n, sep = "")),
                 each = i + 1),
        N2 = rep(sapply(1:(i + 1), function(n) paste(i + 1, "_S", n, sep = "")),
                 time = i),
        Value = as.vector(t(Bratwurst:::nnls_sol(W.list[[as.character(i)]], W.list[[as.character(i + 1)]])))
        )
      df$ID <- paste(df$N1, df$N2)
      return(df)
    }))
    edges_dfList <- split(edges, f = edges$N2)
    edges_vecList <- lapply(edges_dfList, function(current_df){
      norm_vec <- current_df$Value / sum(current_df$Value)
      names(norm_vec) <- current_df$ID
      return(norm_vec)
    })
    edges_vec <- do.call(c, edges_vecList)
    names(edges_vec) <- unlist(lapply(edges_vecList, names))
    edges$rescaled <- edges_vec[as.character(edges$ID)]
    edges <- edges[edges$Value > edges.cutoff, ]
    nodes$y <- Bratwurst:::yNodeCoords(nodes, edges, rank_flag = TRUE,
                           start_coords = c(-0.5, 0.5),
                           edgeWeightColName = "rescaled",
                           scale_fun = function(x){return(x^(1 / 3))})
    edges <- Bratwurst:::reorderEdges(nodes, edges)
    if (color){
      pca <- prcomp(t(do.call(cbind, W.list)))
      pca <- apply(pca$x, 2, function(r) {
        r <- r - min(r)
        return(r / max(r))
      })
      col <- rgb(pca[, 1], pca[, 2], pca[, 3], 0.9)
      nodes$col <- col
    }
    ret <- makeRiver(nodes = nodes, edges = edges)
    return(ret)
  }
  
  
  
environment(my_generateRiverplot) <- asNamespace("Bratwurst")
```


